---
title: "Class 19: Investigating Pertussis"
author: "Wade Ingersoll (PID: A69038080)"
format: pdf
toc: true
---

## Background

Pertussis (a.k.a. Whooping cough) is a highly contagious lung infection caused by the bacteria *B. pertussis*.

The CDC tracks case numbers in the US and makes this data available online
\newpage
> **Q1:** With the help of the R “addin” package `datapasta` assign the CDC pertussis case number data to a data frame called **cdc** and use **ggplot** to make a plot of cases numbers over time.

```{r, include=FALSE}
cdc <- data.frame(
                                 year = c(1922L,1923L,1924L,1925L,
                                          1926L,1927L,1928L,1929L,1930L,1931L,
                                          1932L,1933L,1934L,1935L,1936L,
                                          1937L,1938L,1939L,1940L,1941L,1942L,
                                          1943L,1944L,1945L,1946L,1947L,
                                          1948L,1949L,1950L,1951L,1952L,
                                          1953L,1954L,1955L,1956L,1957L,1958L,
                                          1959L,1960L,1961L,1962L,1963L,
                                          1964L,1965L,1966L,1967L,1968L,1969L,
                                          1970L,1971L,1972L,1973L,1974L,
                                          1975L,1976L,1977L,1978L,1979L,1980L,
                                          1981L,1982L,1983L,1984L,1985L,
                                          1986L,1987L,1988L,1989L,1990L,
                                          1991L,1992L,1993L,1994L,1995L,1996L,
                                          1997L,1998L,1999L,2000L,2001L,
                                          2002L,2003L,2004L,2005L,2006L,2007L,
                                          2008L,2009L,2010L,2011L,2012L,
                                          2013L,2014L,2015L,2016L,2017L,2018L,
                                          2019L,2020L,2021L,2022L,2023L),
         cases = c(107473,164191,165418,152003,
                                          202210,181411,161799,197371,
                                          166914,172559,215343,179135,265269,
                                          180518,147237,214652,227319,103188,
                                          183866,222202,191383,191890,109873,
                                          133792,109860,156517,74715,69479,
                                          120718,68687,45030,37129,60886,
                                          62786,31732,28295,32148,40005,
                                          14809,11468,17749,17135,13005,6799,
                                          7717,9718,4810,3285,4249,3036,
                                          3287,1759,2402,1738,1010,2177,2063,
                                          1623,1730,1248,1895,2463,2276,
                                          3589,4195,2823,3450,4157,4570,
                                          2719,4083,6586,4617,5137,7796,6564,
                                          7405,7298,7867,7580,9771,11647,
                                          25827,25616,15632,10454,13278,
                                          16858,27550,18719,48277,28639,32971,
                                          20762,17972,18975,15609,18617,
                                          6124,2116,3044,7063)
       )
```

```{r}
library(ggplot2)

ggplot(cdc) + 
  aes(year, cases) +
  geom_point() +
  geom_line() +
  labs(x="Year", y="Number of Cases", title = "Pertussis Cases by Year (1922-2019)")
```
\newpage
> **Q2:** Using the ggplot `geom_vline()` function add lines to your previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine (see example in the hint below). What do you notice? Finally, add a line for 2020.

> **Answer:** I noticed that approximately 10 years after switching to the aP vaccine, cases began to rise.

> **Q2.1:** Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

> **Answer:** After introduction of the aP vaccine, the number of cases increased. This could be due to aP's decreased ability to confer lasting immunity compared to wP.

```{r}
ggplot(cdc) + 
  aes(year, cases) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept = 1946, col="blue", linetype="dashed") +
  geom_vline(xintercept = 1996, col="red", linetype="dashed") +
  geom_vline(xintercept = 2020, col="grey", linetype="dashed") +
  labs(x="Year", y="Number of Cases", title = "Pertussis Cases by Year (1922-2019)")
```
\newpage
## The CMI-PB project

The CMI-Pertussis Boost (PB) project focuses on gathering data on this very topic. What is distinct between aP and wP individuals over time when they encounter Pertussis again.

They make their data available via a JSON format returning API. We can read JSON format with the `read_json()` function from the **jsonlite** package with `install.packages("jsonlite")`

```{r}
library(jsonlite)
subject <- read_json("http://cmi-pb.org/api/v5_1/subject", simplifyVector = TRUE)
head(subject)
```

> **Q3:** How many "subjects" (or individuals) are in this dataset?

> **Answer:** 172 (see code below for calculations)

```{r}
nrow(subject)
```
\newpage
> **Q4:** How many aP and wP infancy vaccinated subjects are in the dataset?

> **Answer:** There are 87 aP and 85 wP (see code below for calculations)

```{r}
table(subject$infancy_vac)
```

> **Q5:** How many Male and Female subjects/patients are in the dataset?

> **Answer:** There are 112 females and 60 males (see code below for calculations)

```{r}
table(subject$biological_sex)
```

> **Q6:** What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)?

> **Answer:** See code below for breakdown

```{r}
table(subject$race, subject$biological_sex)
```
\newpage
### Side-Note: Working with dates

Two of the columns of `subject` contain dates in the Year-Month-Day format. Recall from our last mini-project that dates and times can be annoying to work with at the best of times. However, in R we have the excellent lubridate package, which can make life allot easier. Here is a quick example to get you started:

```{r}
library(lubridate)
```

What is today’s date (at the time I am writing this obviously)

```{r}
today()
```

How many days have passed since new year 2000

```{r}
today() - ymd("2000-01-01")
```

What is this in years?

```{r}
time_length( today() - ymd("2000-01-01"),  "years")
```
\newpage
Note that here we are using the `ymd()` function to tell lubridate the format of our particular date and then the `time_length()` function to convert days to years.

> **Q7:** Using this approach determine (i) the average age of wP individuals, (ii) the average age of aP individuals; and (iii) are they significantly different?

>**Answer:** The average age of **wP = 37** and **aP = 28**. These values are statistically significant with a **p-value of 2.372101e-23**. See code below for calculations.

```{r}
# Using ymd() function on the appropriate column of our subject data frame together with today’s date we can calculate the age in days for subjects. We can optionally store this back into our subject data frame for later use:

subject$age <- today() - ymd(subject$year_of_birth)
```

```{r}
library(dplyr)
```

```{r}
# Next we can use dplyr’s filter() function to limit ourselves to a particular subset of subjects to examine the 6 number summary of their age in years:

ap <- subject %>% filter(infancy_vac == "aP")

round( summary( time_length( ap$age, "years" ) ) )
```

```{r}
# wP
wp <- subject %>% filter(infancy_vac == "wP")
round( summary( time_length( wp$age, "years" ) ) )
```

```{r}
# Are they significantly different?
t <- t.test(ap$age, wp$age)
t$p.value
```

> **Q8:** Determine the age of all individuals at time of boost?

> **Answer:** See code below for calculations

```{r}
int <- ymd(subject$date_of_boost) - ymd(subject$year_of_birth)
age_at_boost <- time_length(int, "year")
subject$age_at_boost <- age_at_boost
head(subject)
```
\newpage
> **Q9a:** With the help of a faceted boxplot or histogram (see below), do you think these two groups are significantly different?

> **Answer:** These groups definitely seem different. This observation is supported by a t-test p-value of **2.372101e-23**

```{r}
ggplot(subject) +
  aes(time_length(age, "year"),
      fill=as.factor(infancy_vac)) +
  geom_histogram(show.legend=FALSE) +
  facet_wrap(vars(infancy_vac), nrow=2) +
  xlab("Age in years")

x <- t.test(time_length( wp$age, "years" ),
       time_length( ap$age, "years" ))

x$p.value
```
### Joining multiple tables

Read the specimen and ab_titer tables into R and store the data as `specimen` and `titer` named data frames.

```{r}
# Complete the API URLs...
specimen <- read_json("http://cmi-pb.org/api/v5_1/specimen", simplifyVector = TRUE)
ab_titer <- read_json("http://cmi-pb.org/api/v5_1/plasma_ab_titer", simplifyVector = TRUE)
```

To know whether a given `specimen_id` comes from an aP or wP individual we need to link (a.k.a. “join” or merge) our `specimen` and `subject` data frames. The excellent **dplyr** package (that we have used previously) has a family of `join()` functions that can help us with this common task:

> **Q9b:** Complete the code to join specimen and subject tables to make a new merged data frame containing all specimen records along with their associated subject details:

```{r}
meta <- inner_join(specimen, subject)
dim(meta)
head(meta)
```
\newpage
> **Q10:** Now using the same procedure join meta with titer data so we can further analyze this data in terms of time of visit aP/wP, male/female etc.

```{r}
abdata <- inner_join(ab_titer, meta)
dim(abdata)
```

> **Q11:** How many specimens (i.e. entries in `abdata`) do we have for each `isotype`?

> **Answer:** See code below for answer

```{r}
table(abdata$isotype)
```

> **Q12:** What are the different $dataset values in abdata and what do you notice about the number of rows for the most “recent” dataset?

> **Answer:** For the most recent dataset (2023), the number of rows increased compared to the two previous years. See code below for other $dataset values.

```{r}
table(abdata$dataset)
```
\newpage
## 4. Examine IgG Ab titer levels

Now using our joined/merged/linked `abdata` dataset `filter()` for IgG `isotype`.

```{r}
igg <- abdata %>% filter(isotype == "IgG")
head(igg)
```
\newpage
> **Q13:** Complete the following code to make a summary boxplot of Ab titer levels (MFI) for all antigens:

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen) +
  geom_boxplot() + 
    xlim(0,75) +
  facet_wrap(vars(visit), nrow=2)
```
\newpage
> **Q14:** Antigen levels time-course plot

```{r}
abdata.21 <- abdata %>% filter(dataset == "2021_dataset")

abdata.21 %>% 
  filter(isotype == "IgG",  antigen == "PT") %>%
  ggplot() +
    aes(x=planned_day_relative_to_boost,
        y=MFI_normalised,
        col=infancy_vac,
        group=subject_id) +
    geom_point() +
    geom_line() +
    geom_vline(xintercept=0, linetype="dashed") +
    geom_vline(xintercept=14, linetype="dashed") +
  labs(title="2021 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)")
```
\newpage
> **Q14** continued

```{r}
igg <- igg %>% filter(visit %in% 1:8)
ggplot(igg) +
  aes(MFI_normalised, antigen, col=infancy_vac) +
  geom_boxplot() +
  facet_wrap(~visit, nrow=2)
```

