---
title: "Class 8: ML Mini Project (Breast Cancer Analysis)"
author: "Wade Ingersoll (PID: A69038080)"
format: pdf
toc: true
---

## Background

The goal of today's mini-project is to explore a complete analysis using the unsupervised learning techniques covered in the last class. We will extend what we learned by combining PCA as a preprocessing step to clustering using data that consist of measurements of cell nuclei of human breast massess.

The data itself comes from the Wisconsin Breast Cancer Diagnostic Data Set first reported by K. P. Benne and O. L. Mangasarian: “Robust Linear Programming Discrimination of Two Linearly Inseparable Sets”.

Values in this data set describe characteristics of the cell nuclei present in digitized images of a fine needle aspiration (FNA) of a breast mass.

## Data import

The data is available as a CSV from class website:

```{r}
wisc.df <- read.csv("WisconsinCancer (2).csv", row.names=1)
```

Make sure we do not include sample id or diagnosis columns in the data that we analyze below.

```{r}
diagnosis <- as.factor(wisc.df$diagnosis)
wisc.data <- wisc.df[, -1]
dim(wisc.data)
```

## Exploratory data analysis

> Q1. How many observations are in this dataset?

There are `r nrow(wisc.data)` observations/samples/patients in the data set.
I used nrow(wisc.data) to find that number.

> Q2. How many of the observations have a malignant diagnosis?

There are 212; see code below (two ways to determine)

```{r}
sum(wisc.df$diagnosis == "M")
```

```{r}
table(wisc.df$diagnosis)
```


> Q3. How many variables/features in the data are suffixed with _mean?

There are 10; see code below

```{r}
#colnames(wisc.data)
length( grep("_mean", colnames(wisc.data)) )
```

Easier to read: 

```{r}
n <- colnames(wisc.data)
inds <- grep("_mean", n)
length(inds)
```

