---
title: "Class05: Data Visualization with ggplot"
author: "Wade Ingersoll (PID: A69038080)"
format: pdf
---

Today we are playing with plotting and graphics in R.

There are lots of ways to make cool figures in R.
There is "base" R graphics (`plot()`, `hist()`, `boxplot()`, etc.)

There is also add-on packages, like **ggplot**.

```{r}
# View the top 6 rows of 'cars' dataset
head(cars)
```
Let's plot this with "base" R:

```{r}
# Plot 'cars' with base R
plot(cars)
```

```{r}
# View top 6 rows of 'mtcars' dataset
head(mtcars)
```

Let's plot `mpg` vs `disp`

```{r}
plot(mtcars$mpg, mtcars$disp, pch=16)
```

```{r}
hist(mtcars$mpg)
```

## GGPLOT

The main function in the ggplot2 package is ``ggplot()``. First, I need to install the **ggplot2** package. I can install any package with the function `install.packages()`

> **N.B.** I never want to run `install.packages()` in my quarto source document!!

```{r}
# In the console, run install.packages("ggplot2"). This only needs to be done once!
# Load ggplot2
library(ggplot2)

# Plot cars dataset with ggplot and add geom_point() to generate scatterplot
ggplot(cars) +
  aes(x=speed, y=dist) +
  geom_point()
```

```{r}
ggplot(cars) +
  aes(speed) +
  geom_histogram()
```

Every ggplot needs at least 3 things: 

- The **data** (given with `ggplot(cars)`)
- The **aesthetic** mapping (given with `aes()`)
- The **geom** (given by `geom_point()`)

> For simple canned graphs "base" R is nearly always faster (in both writing the code and executing it)

### Adding more layers ###

Let's add a line and a title, subtitle and caption as well as custom axis labels

```{r}
ggplot(cars) +
  aes(x=speed, y=dist) +
  geom_point() +
  geom_line()
```

```{r}
ggplot(cars) +
  aes(x=speed, y=dist) +
  geom_point() +
  geom_smooth(method="lm", se=FALSE) +
  labs(title = "GGPLOT PRACTICE", 
       subtitle = "Wed Oct 15, 2025", 
       x = "Speed (mph)", 
       y = "Distance (ft)",
       caption = "BGGN213") +
  theme_bw()
```

## Plot some expression data

Read data file from online URL

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

> Q1. How many genes are in this dataset?

There are `r nrow(genes)` in this dataset

> Q2. How many "up" regulated genes are there?

```{r}
sum(genes$State == "up")
```

There are `r sum(genes$State == "up")` UP genes

```{r}
table(genes$State)
```

```{r}
ggplot(genes) +
  aes(x=Condition1, y=Condition2, col=State) +
  geom_point()
```


```{r}
p <- ggplot(genes) +
  aes(x=Condition1, y=Condition2, col=State) +
  geom_point() +
  scale_color_manual(values=c("blue", "grey", "red"))
```

```{r}
p + labs(title="hello")
```

Example of adding labels

```{r}
library(ggrepel)

ggplot(genes) +
  aes(x=Condition1, 
      y=Condition2, 
      col=State,
      label=Gene) +
  geom_point() +
  scale_color_manual(values=c("blue", "grey", "red")) +
  geom_text_repel(max.overlaps = 200) +
  theme_bw() +
  labs(title="Example plot with labs setting")
```

## Going further

Playing with some different layers and the gapminder dataset...

```{r}
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
```

```{r}
head(gapminder)
```

```{r}
tail(gapminder)
```

Plot gdpPercap (x) vs lifeExp (y)

```{r}
ggplot(gapminder) +
  aes(x=gdpPercap, y=lifeExp, col=continent) +
  geom_point(alpha=0.3, shape=25) +
  facet_wrap(~continent) +
  theme_bw() +
  scale_size_area(max_size = 1)
```
