---
title: "Class 16: Analyzing Sequencing Data in the Cloud"
author: "Wade Ingersoll (PID: 69038080)"
format: pdf
---

# Downstream Analysis

Back on our laptop we can now use R and Bioconductor tools to further explore this large scale dataset.

For example there is an R function called `tximport()` in the **tximport** package, which enables straightforward import of Kallisto results

With each sample having its own directory containing the Kallisto output, we can import the transcript count estimates into R using:

```{r}
# BiocManager::install("tximport")
# BiocManager::install("rhdf5")
```

```{r}
library(tximport)
library(rhdf5)

# setup the folder and filenames to read
folders <- dir(pattern="SRR21568*")
samples <- sub("_quant", "", folders)
files <- file.path( folders, "abundance.h5" )
names(files) <- samples

txi.kallisto <- tximport(files, type = "kallisto", txOut = TRUE)
```
```{r}
head(txi.kallisto$counts)
```

We now have our estimated transcript counts for each sample in R. We can see how many transcripts we have for each sample:

```{r}
colSums(txi.kallisto$counts)
```

And how many transcripts are detected in at least one sample:

```{r}
sum(rowSums(txi.kallisto$counts)>0)
```

Before subsequent analysis, we might want to filter out those annotated transcripts with no reads:

```{r}
to.keep <- rowSums(txi.kallisto$counts) > 0
kset.nonzero <- txi.kallisto$counts[to.keep,]
```

And those with no change over the samples:

```{r}
keep2 <- apply(kset.nonzero,1,sd)>0
x <- kset.nonzero[keep2,]
```
\newpage
## Principal Component Analysis

We can now apply any exploratory analysis technique to this counts matrix. As an example, we will perform a PCA of the transcriptomic profiles of these samples.

Now we compute the principal components, centering and scaling each transcript’s measured levels so that each feature contributes equally to the PCA:

```{r}
pca <- prcomp(t(x), scale=TRUE)
```

```{r}
summary(pca)
```

Now we can use the first two principal components as a co-ordinate system for visualizing the summarized transcriptomic profiles of each sample:

```{r}
plot(pca$x[,1], pca$x[,2],
     col=c("blue","blue","red","red"),
     xlab="PC1", ylab="PC2", pch=16)
```
\newpage
> Q. Use ggplot to make a similar figure of PC1 vs PC2 and a seperate figure PC1 vs PC3 and PC2 vs PC3.

**PC1 vs PC2:**

```{r}
library(ggplot2)
library(ggrepel)

# Make metadata object for the samples
colData <- data.frame(condition = factor(rep(c("control", "treatment"), each = 2)))
rownames(colData) <- colnames(txi.kallisto$counts)

# Make the data.frame for ggplot 
y <- as.data.frame(pca$x)
y$Condition <- as.factor(colData$condition)

ggplot(y) +
  aes(PC1, PC2, col=Condition) +
  geom_point() +
  geom_text_repel(label=rownames(y)) +
  theme_bw() +
  ggtitle("PC1 vs PC2")
```
\newpage
**PC1 vs PC3:**

```{r}
library(ggplot2)
library(ggrepel)

# Make metadata object for the samples
colData <- data.frame(condition = factor(rep(c("control", "treatment"), each = 2)))
rownames(colData) <- colnames(txi.kallisto$counts)

# Make the data.frame for ggplot 
y <- as.data.frame(pca$x)
y$Condition <- as.factor(colData$condition)

ggplot(y) +
  aes(PC1, PC3, col=Condition) +
  geom_point() +
  geom_text_repel(label=rownames(y)) +
  theme_bw() +
  ggtitle("PC1 vs PC3")
```
\newpage
**PC2 vs PC3:**

```{r}
library(ggplot2)
library(ggrepel)

# Make metadata object for the samples
colData <- data.frame(condition = factor(rep(c("control", "treatment"), each = 2)))
rownames(colData) <- colnames(txi.kallisto$counts)

# Make the data.frame for ggplot 
y <- as.data.frame(pca$x)
y$Condition <- as.factor(colData$condition)

ggplot(y) +
  aes(PC2, PC3, col=Condition) +
  geom_point() +
  geom_text_repel(label=rownames(y)) +
  theme_bw() +
  ggtitle("PC2 vs PC3")
```

The plot makes it clear that PC1 separates the two control samples (SRR2156848 and SRR2156849) from the two enhancer-targeting CRISPR-Cas9 samples (SRR2156850 and SRR2156851). PC2 separates the two control samples from each other, and PC3 separates the two enhancer-targeting CRISPR samples from each other. This could be investigated further to see which genes result in these separation patterns. It is also at least slightly reassuring, implying that there are considerable differences between the treated and control samples.

# Differential Gene Expression

We can use **DESeq2** to complete the differential-expression analysis that we are already familiar with:

An example of creating a DESeqDataSet for use with DESeq2:

```{r}
library(DESeq2)
```

```{r}
sampleTable <- data.frame(condition = factor(rep(c("control", "treatment"), each = 2)))
rownames(sampleTable) <- colnames(txi.kallisto$counts)
```

```{r}
dds <- DESeqDataSetFromTximport(txi.kallisto,
                                sampleTable, 
                                ~condition)
```
```{r}
# dds is now ready for DESeq() see our previous classes on this
```

```{r}
dds <- DESeq(dds)
```
```{r}
res <- results(dds)
head(res)
```
These results could go on to be visualized and subjected to pathway analysis etc. as we have done in previous classes.

```{r}
plot(res$log2FoldChange, log='xy')
```

